Spike-and-Slab LASSO Generalized Additive Models and Scalable Algorithms for High-Dimensional Data Analysis

### 2022-05-16 Manuscript/SS_GAM.tex
update manuscript
Replace "smoothing function" with "smooth function"

\abstract{There are proposals that extend the classical generalized
additive models (GAMs) to accommodate high-dimensional data (\(p>>n\))
using group sparse regularization. However, the sparse regularization
may induce excess shrinkage when estimating smoothing functions,
damaging predictive performance. Moreover, most of these GAMs consider
an ``all-in-all-out'' approach for functional selection, rendering them
difficult to answer if nonlinear effects are necessary. While some
Bayesian models can address these shortcomings, using Markov chain Monte
Carlo algorithms for model fitting creates a new challenge, scalability.
Hence, we propose Bayesian hierarchical generalized additive models as a
solution: we consider the smoothing penalty for proper shrinkage of
curve interpolation via reparameterization. A novel two-part
spike-and-slab LASSO prior for smoothing functions is developed to
address the sparsity of signals while providing extra flexibility to
select the linear or nonlinear components of smoothing functions. A
scalable and deterministic algorithm, EM-Coordinate Descent, is
implemented in an open-source R package BHAM. Simulation studies and
metabolomics data analyses demonstrate improved predictive and
computational performance against state-of-the-art models. Functional
selection performance suggests trade-offs exist regarding the effect
hierarchy assumption.}

@@

\textcolor{red}{For their straightforward interpretation and flexibility, nonparametric regression models serve as great alternatives to the "black-box" models in prediction and variable selection}.
Among those, generalized additive models (GAMs), proposed in the seminal
work of Hastie and Tibshirani \citep{Hastie1987}, grew to be one of the
most popular modeling tools. In a GAM, the response variable, which is
assumed to follow some exponential family distribution, can be modeled
with the summation of smoothing functions. Nevertheless, the classical
GAMs cannot fulfill the increasing analytic demands for high-dimensional
data analysis.

There exist some proposals to generalize the classical GAM to
accommodate high-dimensional applications. The regularized models,
branching out from group regularized linear models, are used to fit GAMs
by accounting for the structure introduced when expanding smoothing
functions. Ravikumar et al. \citep{Ravikumar2009} extended the group
LASSO \citep{Yuan2006} to additive models (AMs); Huang et al.
\citep{Huang2010} further developed adaptive group LASSO for additive
models; Wang et al. \citep{Wang2007} and Xue \citep{Xue2009}
respectively applied group SCAD penalty \citep{Fan2001} to additive
models. Bayesian hierarchical models are also used in the context of
high-dimensional additive models, particularly within the spike-and-slab
literature. Various group spike-and-slab priors \citep{Xu2015, Yang2020}
combining with computationally intensive Markov chain Monte Carlo (MCMC)
algorithms are proposed, where the application on AMs is treated as a
special case. Bai and co-authors\citep{Bai2020} were the first to apply
group spike-and-slab LASSO prior to Gaussian AMs using a fast
optimization algorithm and further generalized the framework to GAMs
\citep{Bai2021}. Focus on addressing the sparsity, these methods can
excessively penalize the bases of a smoothing function and produce
inaccurate predictions, particularly when complex signals are assumed
and large numbers of knots are used. \citep{Scheipl2013} In addition,
these methods adopt an `all-in-all-out' strategy, i.e.~either including
or excluding the variable completely, rendering no space for bi-level
selection. Scheipl et al. \citep{Scheipl2012} proposed a spike-and-slab
structure prior that addresses the bi-level selection. But the model
fitting relies on computationally intensive MCMC algorithms and creates
scalability concerns. Developing a fast, flexible and accurate
generalized additive model framework would be of special interest.

\textcolor{red}{We propose a novel Bayesian hierarchical generalized additive model (BHAM) for outcome prediction in the context of high-dimensional data analysis. Specifically, we incorporate smoothing penalties, derived from the smoothing spline literature \cite{Wood2017}, via reparameterization of smoothing functions to avoid excessive shrinkage on the bases. Smoothing penalties were also previously used in the spike-and-slab GAM \cite{Scheipl2012} and the sparsity-smoothness penalty \cite{Meier2009}. We then impose a new two-part spike-and-slab LASSO prior to address the signal sparsity. In addition, a scalable optimization-based algorithm, EM-Coordinate Descent (EM-CD) algorithm, is developed. While the primary focus of this model is to improve prediction, the proposed model also provides utility in functional selection. Notably, the two-part prior that follows the effect hierarchy principle motivates a bi-level selection, rendering one of three possibilities for each predictor: no effect, only linear effect, or linear and nonlinear effects. The proposed model is implemented in a publicly available R package \texttt{BHAM} via \url{https://github.com/boyiguo1/BHAM}.}

The proposed framework, BHAM, differs from previous spike-and-slab based
GAMs, i.e., the spike-and-slab GAM \citep{Scheipl2012} and the sparse
Bayesian GAM (SB-GAM) \citep{Bai2021}, in three ways. Firstly, the
proposed prior for smoothing functions is a spike-and-slab LASSO type
prior using independent mixture double exponential distribution,
compared to spike-and-slab GAM that uses normal-mixture-of-inverse gamma
prior. Spike-and-slab LASSO priors provide computational convenience
during model fitting by using optimization algorithms instead of
intensive sampling algorithms. They make fitting high-dimensional models
more feasible without sacrificing performance in prediction and variable
selection. Secondly, SB-GAM uses a group spike-and-slab LASSO prior with
an EM-CD algorithm to fit the model. While both methods use the
combination of expectation maximization algorithm and coordinate descent
algorithm, there are subtle differences in the implementation due to the
difference in prior specification. The proposed model sets up
independent priors among basis coefficients after the reparameterization
step, which provides some advantage in computation. Lastly, the proposed
model addresses the incapability of bi-level selection in SB-GAM.

@@

\textcolor{red}{We assume the response variable, $Y$, follows an exponential family distribution with density function $f(y)$, mean $\mu$ and dispersion parameter $\phi$. The mean of the response variable can be modeled as the summation of smoothing functions, $B_j(\cdot), j = 1, \dots, p$, of a given $p$-dimensional vector of predictors $\boldsymbol{x}$, written as 
\begin{equation}\label{eq:gam}
 E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation}
where $g^{-1}(\cdot)$ is the inverse of a monotonic link function. Given $n$ data points $\{y_i, \boldsymbol{x}_i\}^n_{i=1}$,}
the data distribution is expressed as \begin{equation}
f(\boldsymbol{Y} = \boldsymbol{y}| \boldsymbol{\beta}, \phi) = \prod\limits^n_{i=1}f( Y = y_i|\boldsymbol{\beta}, \phi).\nonumber
\end{equation}

The basis function matrix, i.e.~the design matrix derived from the
smoothing function \(B_j(x_j)\), is denoted \(\boldsymbol{X}_j\) for the
variable \(x_j\). The dimension of the design matrix depends on the
choice of the smoothing function, and is denoted as \(K_j\) for \(x_j\).
\(\boldsymbol{\beta}_j\) denotes the basis function coefficients for the
\(j\)th variable such that
\(B_j(x_j) = \boldsymbol{\beta}_j^T \boldsymbol{X}_j\). With slight
abuse of notation, we denote vectors and matrices in bold fonts
\(\boldsymbol{\beta}, \boldsymbol{X}\) with conformable dimensions,
where scalar and random variables are denoted in unbold fonts
\(\beta, X\). The matrix transposing operation is denoted with a
superscript \(^T\).
\textcolor{red}{To note, the proposed model can include parametric forms of variables in the model, and hence considers general linear models and semiparametric regression models as special cases.}

\subsection{Smoothing Function Reparameterization}

@@

Marra and Wood \citep{Marra2011} proposed a reparameterization procedure
\textcolor{red}{to factor the smoothing penalty into the design matrix of each smoothing function}.
Given that the smoothing penalty matrix \(\boldsymbol{S}_j\) is
symmetric and positive semi-definite for the univariate smoothing
functions, we eigendecompose the penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\) ,
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
elements of zeros on the diagonal, where the zeros are associated with
the linear space of the smoothing function. For the most popular
smoothing function, cubic splines, the dimension of the linear space is
one. Hereafter, we focus on discussing a uni-dimensional linear space
for simplicity; however, it generalizes easily to the cases where the
linear space is multidimensional. We further write the orthonormal
matrix
\(\boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{U}^0 : \boldsymbol{U}^{*}\end{bmatrix}\)
containing the eigenvectors as columns in the corresponding order to
\(\boldsymbol{D}\). That is, \(\boldsymbol{U}\) contains the
eigenvectors \(U^0\) with zero eigenvalues for the linear space and
\(\boldsymbol{U}^{*}\) contains the eigenvectors (as columns) for the
non-zero eigenvalues, i.e.~the nonlinear space. We multiply the basis
function matrix \(\boldsymbol{X}\) with the orthonormal matrix
\(\boldsymbol{U}\) for the new design matrix
\({\boldsymbol{X}}^\text{repa}= \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} X^0 : \boldsymbol{X}^{*} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{*}\) by the
non-zero eigenvalues of \(\boldsymbol{D}\) such that the new basis
function matrix \(\boldsymbol{X}^\ast\) can receive a uniform penalty on
each of its dimensions. With slight abuse of the notation, we drop the
superscript \(^\text{repa}\) and denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} X_j^0 : \boldsymbol{X}_j^{*} \end{bmatrix}\)
as the basis function matrix for the \(j\)th variable after the
reparameterization. A spline function can be expressed in the matrix
form \[
B_j(x_j) = B_j^0(x_j) + B_j^*(x_j) = \beta_j X^0_j + \boldsymbol{\beta_j^*}^T \boldsymbol{X}_j^*,
\] and the generalized additive model in Equation (\ref{eq:gam}) now is
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[\beta_0 + \sum\limits^p_{j=1} (\beta_j X^0_j + {\boldsymbol{\beta}_j^*}^T \boldsymbol{X}_j^*)\right],
\end{equation} where the coefficients
\(\boldsymbol{\beta}_j \equiv \begin{bmatrix} \beta_j : \boldsymbol{\beta}^*_j \end{bmatrix}\)
is an augmentation of the coefficient scalar \(\beta_j\) of linear space
and the coefficient vector \(\boldsymbol{\beta}^*_j\) of nonlinear
space.

@@

We introduce a novel SSL-based prior for smoothing functions in GAMs.
Given the reparameterized design matrix
\(\boldsymbol{X}_j = \begin{bmatrix} X^0_j : \boldsymbol{X}_j^*\end{bmatrix}\)
for the \(j\)th variable, we impose a two-part SSL prior to the
coefficients
\(\boldsymbol{\beta}_j = \begin{bmatrix} \beta_j : \boldsymbol{\beta}_j^*\end{bmatrix}\).
Specifically, the linear space coefficient has an SSL prior and the
nonlinear space coefficients share a group SSL prior,
\begin{align}\label{eq:bham_ssl}
  \beta_{j} | \gamma_{j},s_0,s_1 &\sim DE(0,(1-\gamma_{j}) s_0 + \gamma_{j} s_1) \nonumber \\
  \beta^*_{jk} | \gamma^*_{j},s_0,s_1 &\overset{\text{iid}}{\sim}DE(0,(1-\gamma^*_{j}) s_0 + \gamma^*_{j} s_1), k=1,\dots, K_j
\end{align} where \(\gamma_{j}\in\{0,1\}\) and
\(\gamma^*_{j}\in \{0,1\}\) are two latent indicator variables,
indicating if the model includes the linear effect and the nonlinear
effect of the \(j\)th variable respectively. \(s_0\) and \(s_1\) are
scale parameters, assuming \(0 < s_0 < s_1\) and given. These scale
parameters \(s_0\) and \(s_1\) can be treated as tuning parameters and
optimized via cross-validation, discussed in Section \ref{sec:tune}.

@@

\textcolor{red}{To allow the shrinkage to self-adapt to the sparsity and smoothing pattern of the data, }we
further specify the parameter \(\theta_j\) follows a beta distribution
with given shape parameters \(a\) and \(b\), \[
\theta_j \sim Beta(a, b).
\] The beta distribution is a conjugate prior to the binomial
distribution and hence provides some computation convenience.
\textcolor{red}{Having a prior distribution of $\theta_j$ enables the proposed prior to inherit the selective shrinkage property and self-adaptivity \cite{Bai2021Review} from the classical SSL prior. In other words, when a smoothing function is significant, the coefficients of the smoothing function escape the overall shrinkage and produce a more accurate estimate, particularly in pair with the smoothing penalty implicitly addressed via the reparameterization. Meanwhile, the hyper prior encourages information borrowing across coordinates and hence automatic adjust for different levels of sparsity.}
Hereafter, we refer to the Bayesian hierarchical generalized additive
models with the two-part spike-and-slab LASSO prior as BHAM, and
visually presented in Figure \ref{fig:SSprior}.

@@

In this section, we compare the performance of the proposed model to
\textcolor{red}{six alternative models: linear LASSO models,} component
selection and smoothing operator (COSSO) \citep{Zhang2006GAM}, adaptive
COSSO \citep{Storlie2011}, generalized additive models with automatic
smoothing (referred to as \textit{mgcv} hereafter)\citep{Wood2011},
\textcolor{red}{spike-and-slab GAM \cite{Scheipl2012}}, and SB-GAM
\citep{Bai2021}.
\textcolor{red}{We use linear LASSO models as the benchmark, examining the performance when the linearity assumption doesn't hold}.
COSSO is one of the earliest smoothing spline models that consider
sparsity-smoothness penalty. Adaptive COSSO improved upon COSSO by using
adaptive weight for penalties such that the penalty of each functional
component is different for extra flexibility. \textit{mgcv} is one of
the most popular models for nonlinear effect interpolation and
prediction. Nevertheless,
\textcolor{red}{mgcv doesn't support analyses when the number of parameters exceeds the sample size.}
\textcolor{red}{Spike-and-slab GAM employs a spike-and-slab prior for GAM and uses an MCMC algorithm for model fitting.}
SB-GAM is the first spike-and-slab LASSO GAM. We implement linear LASSO
model with R package \texttt{glmnet 4.1-2}, COSSO and adaptive COSSO
with R package \texttt{cosso 2.1-1}, generalized additive models with
automatic smoothing with R package \texttt{mgcv 1.8-31}, spike-and-slab
GAM with R package \texttt{spikeSlabGAM 1.1-15}, and SB-GAM with R
package \texttt{sparseGAM 1.0}. COSSO models and SB-GAM do not provide
flexibility to define smoothing functions, and hence use the default
choices; mgcv, spikeSlabGAM and the proposed model allow customized
smoothing functions and we choose the cubic regression spline. We
control the dimensionality of each smoothing function, 10 bases, for all
different choices of smoothing functions. We use a 5-fold CV with the
default selection criteria to select the final model for linear LASSO
model, COSSO models, SB-GAM and the proposed model. Twenty default
candidates of tuning parameters (\(s_0\) in BHAM, \(\lambda_0\) in
SB-GAM) are examined for SB-GAM and the proposed model that allow user
specification of tuning candidates. All computation was conducted on a
high-performance 64-bit Linux platform with 48 cores of 2.70GHz
eight-core Intel Xeon E5-2680 processors and 24G of RAM per core and R
3.6.2 \citep{R}.

@@

\textcolor{red}{In the paper, we described a novel high-dimensional generalized additive model with Bayesian hierarchical prior for the purpose of predictive modeing. In particular, we introduce a two-part spike-and-slab LASSO prior for reparameterized smoothing functions and derive a scalable EM-CD algorithm for model fitting. The proposed model provides a new angle to address the excess shrinkage of smoothing functions that is commonly vulnerable to previous regularized high-dimensional GAMs, and hence improves the predictive performance. Th EM-CD algorithm, extended from previous spike-and-slab LASSO models, provides a computationally efficient alternative to the computational prohibitive MCMC algorithms, enhancing the scalability of spike-and-slab models. In addition, the two-part prior motivates the bi-level selection of predictors, selection of linear and nonlinear components. In the simulation study and real-data analyses, the proposed model demonstrates improvement in prediction and computational advantage when compared to the state-of-the-art models. When serving the purpose of variable selection, trade-offs exist among methods of comparison.}
We implement the proposed model in an open-source R package
\texttt{BHAM}, deposited at \url{https://github.com/boyiguo1/BHAM}. To
maximize the flexibility of smoothing function specification, we deploy
the same programming grammar as in the state-of-the-art package
\texttt{mgcv}, in contrast to previous tools where smoothing functions
are limited to the default ones. Ancillary functions are provided for
model specification in high-dimensional settings, curve plotting and
functional selection.

The proposed model shares many commonalities with the SB-GAM
\citep{Bai2021}, which is independently developed around the same time
as the proposed work. Both frameworks emphasize computational efficiency
by deploying group spike-and-slab LASSO type priors and
optimization-based scalable algorithms. Bai provides the theoretical
proof for the consistency of variable selection using group
spike-and-slab LASSO prior.
\textcolor{red}{The proposed model focuses on improving prediction performance for high-dimensional GAM, with the capability of bi-level selection. Moreover, the proposed model can easily generalize to other families of priors and smoothing functions if desired. Not focused in this manuscript, the generalization is described in the Supporting Information.}

>>>>

\abstract{There are proposals that extend the classical generalized
additive models (GAMs) to accommodate high-dimensional data (\(p>>n\))
using group sparse regularization. However, the sparse regularization
may induce excess shrinkage when estimating smooth functions, damaging
predictive performance. Moreover, most of these GAMs consider an
``all-in-all-out'' approach for functional selection, rendering them
difficult to answer if nonlinear effects are necessary. While some
Bayesian models can address these shortcomings, using Markov chain Monte
Carlo algorithms for model fitting creates a new challenge, scalability.
Hence, we propose Bayesian hierarchical generalized additive models as a
solution: we consider the smoothing penalty for proper shrinkage of
curve interpolation via reparameterization. A novel two-part
spike-and-slab LASSO prior for smooth functions is developed to address
the sparsity of signals while providing extra flexibility to select the
linear or nonlinear components of smooth functions. A scalable and
deterministic algorithm, EM-Coordinate Descent, is implemented in an
open-source R package BHAM. Simulation studies and metabolomics data
analyses demonstrate improved predictive and computational performance
against state-of-the-art models. Functional selection performance
suggests trade-offs exist regarding the effect hierarchy assumption.}

@@

\textcolor{red}{For their straightforward interpretation and flexibility, nonparametric regression models serve as great alternatives to the "black-box" models in prediction and variable selection}.
Among those, generalized additive models (GAMs), proposed in the seminal
work of Hastie and Tibshirani \citep{Hastie1987}, grew to be one of the
most popular modeling tools. In a GAM, the response variable, which is
assumed to follow some exponential family distribution, can be modeled
with the summation of smooth functions. Nevertheless, the classical GAMs
cannot fulfill the increasing analytic demands for high-dimensional data
analysis.

There exist some proposals to generalize the classical GAM to
accommodate high-dimensional applications. The regularized models,
branching out from group regularized linear models, are used to fit GAMs
by accounting for the structure introduced when expanding smooth
functions. Ravikumar et al. \citep{Ravikumar2009} extended the group
LASSO \citep{Yuan2006} to additive models (AMs); Huang et al.
\citep{Huang2010} further developed adaptive group LASSO for additive
models; Wang et al. \citep{Wang2007} and Xue \citep{Xue2009}
respectively applied group SCAD penalty \citep{Fan2001} to additive
models. Bayesian hierarchical models are also used in the context of
high-dimensional additive models, particularly within the spike-and-slab
literature. Various group spike-and-slab priors \citep{Xu2015, Yang2020}
combining with computationally intensive Markov chain Monte Carlo (MCMC)
algorithms are proposed, where the application on AMs is treated as a
special case. Bai and co-authors\citep{Bai2020} were the first to apply
group spike-and-slab LASSO prior to Gaussian AMs using a fast
optimization algorithm and further generalized the framework to GAMs
\citep{Bai2021}. Focus on addressing the sparsity, these methods can
excessively penalize the bases of a smooth function and produce
inaccurate predictions, particularly when complex signals are assumed
and large numbers of knots are used. \citep{Scheipl2013} In addition,
these methods adopt an `all-in-all-out' strategy, i.e.~either including
or excluding the variable completely, rendering no space for bi-level
selection. Scheipl et al. \citep{Scheipl2012} proposed a spike-and-slab
structure prior that addresses the bi-level selection. But the model
fitting relies on computationally intensive MCMC algorithms and creates
scalability concerns. Developing a fast, flexible and accurate
generalized additive model framework would be of special interest.

\textcolor{red}{We propose a novel Bayesian hierarchical generalized additive model (BHAM) for outcome prediction in the context of high-dimensional data analysis. Specifically, we incorporate smoothing penalties, derived from the smoothing spline literature \cite{Wood2017}, via reparameterization of smooth functions to avoid excessive shrinkage on the bases. Smoothing penalties were also previously used in the spike-and-slab GAM \cite{Scheipl2012} and the sparsity-smoothness penalty \cite{Meier2009}. We then impose a new two-part spike-and-slab LASSO prior to address the signal sparsity. In addition, a scalable optimization-based algorithm, EM-Coordinate Descent (EM-CD) algorithm, is developed. While the primary focus of this model is to improve prediction, the proposed model also provides utility in functional selection. Notably, the two-part prior that follows the effect hierarchy principle motivates a bi-level selection, rendering one of three possibilities for each predictor: no effect, only linear effect, or linear and nonlinear effects. The proposed model is implemented in a publicly available R package \texttt{BHAM} via \url{https://github.com/boyiguo1/BHAM}.}

The proposed framework, BHAM, differs from previous spike-and-slab based
GAMs, i.e., the spike-and-slab GAM \citep{Scheipl2012} and the sparse
Bayesian GAM (SB-GAM) \citep{Bai2021}, in three ways. Firstly, the
proposed prior for smooth functions is a spike-and-slab LASSO type prior
using independent mixture double exponential distribution, compared to
spike-and-slab GAM that uses normal-mixture-of-inverse gamma prior.
Spike-and-slab LASSO priors provide computational convenience during
model fitting by using optimization algorithms instead of intensive
sampling algorithms. They make fitting high-dimensional models more
feasible without sacrificing performance in prediction and variable
selection. Secondly, SB-GAM uses a group spike-and-slab LASSO prior with
an EM-CD algorithm to fit the model. While both methods use the
combination of expectation maximization algorithm and coordinate descent
algorithm, there are subtle differences in the implementation due to the
difference in prior specification. The proposed model sets up
independent priors among basis coefficients after the reparameterization
step, which provides some advantage in computation. Lastly, the proposed
model addresses the incapability of bi-level selection in SB-GAM.

@@

\textcolor{red}{We assume the response variable, $Y$, follows an exponential family distribution with density function $f(y)$, mean $\mu$ and dispersion parameter $\phi$. The mean of the response variable can be modeled as the summation of smooth functions, $B_j(\cdot), j = 1, \dots, p$, of a given $p$-dimensional vector of predictors $\boldsymbol{x}$, written as 
\begin{equation}\label{eq:gam}
 E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1}B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j),
\end{equation}
where $g^{-1}(\cdot)$ is the inverse of a monotonic link function. Given $n$ data points $\{y_i, \boldsymbol{x}_i\}^n_{i=1}$,}
the data distribution is expressed as \begin{equation}
f(\boldsymbol{Y} = \boldsymbol{y}| \boldsymbol{\beta}, \phi) = \prod\limits^n_{i=1}f( Y = y_i|\boldsymbol{\beta}, \phi).\nonumber
\end{equation}

The basis function matrix, i.e.~the design matrix derived from the
smooth function \(B_j(x_j)\), is denoted \(\boldsymbol{X}_j\) for the
variable \(x_j\). The dimension of the design matrix depends on the
choice of the smooth function, and is denoted as \(K_j\) for \(x_j\).
\(\boldsymbol{\beta}_j\) denotes the basis function coefficients for the
\(j\)th variable such that
\(B_j(x_j) = \boldsymbol{\beta}_j^T \boldsymbol{X}_j\). With slight
abuse of notation, we denote vectors and matrices in bold fonts
\(\boldsymbol{\beta}, \boldsymbol{X}\) with conformable dimensions,
where scalar and random variables are denoted in unbold fonts
\(\beta, X\). The matrix transposing operation is denoted with a
superscript \(^T\).
\textcolor{red}{To note, the proposed model can include parametric forms of variables in the model, and hence considers general linear models and semiparametric regression models as special cases.}

\subsection{smooth function Reparameterization}

@@

Marra and Wood \citep{Marra2011} proposed a reparameterization procedure
\textcolor{red}{to factor the smoothing penalty into the design matrix of each smooth function}.
Given that the smoothing penalty matrix \(\boldsymbol{S}_j\) is
symmetric and positive semi-definite for the univariate smooth
functions, we eigendecompose the penalty matrix
\(\boldsymbol{S} = \boldsymbol{U} \boldsymbol{D} \boldsymbol{U}^T\) ,
where the matrix \(\boldsymbol{D}\) is diagonal with the eigenvalues
arranged in the ascending order. To note, \(\boldsymbol{D}\) can contain
elements of zeros on the diagonal, where the zeros are associated with
the linear space of the smooth function. For the most popular smooth
function, cubic splines, the dimension of the linear space is one.
Hereafter, we focus on discussing a uni-dimensional linear space for
simplicity; however, it generalizes easily to the cases where the linear
space is multidimensional. We further write the orthonormal matrix
\(\boldsymbol{U} \equiv \begin{bmatrix} \boldsymbol{U}^0 : \boldsymbol{U}^{*}\end{bmatrix}\)
containing the eigenvectors as columns in the corresponding order to
\(\boldsymbol{D}\). That is, \(\boldsymbol{U}\) contains the
eigenvectors \(U^0\) with zero eigenvalues for the linear space and
\(\boldsymbol{U}^{*}\) contains the eigenvectors (as columns) for the
non-zero eigenvalues, i.e.~the nonlinear space. We multiply the basis
function matrix \(\boldsymbol{X}\) with the orthonormal matrix
\(\boldsymbol{U}\) for the new design matrix
\({\boldsymbol{X}}^\text{repa}= \boldsymbol{X} \boldsymbol{U} \equiv \begin{bmatrix} X^0 : \boldsymbol{X}^{*} \end{bmatrix}\).
An additional scaling step is imposed on \(\boldsymbol{X}^{*}\) by the
non-zero eigenvalues of \(\boldsymbol{D}\) such that the new basis
function matrix \(\boldsymbol{X}^\ast\) can receive a uniform penalty on
each of its dimensions. With slight abuse of the notation, we drop the
superscript \(^\text{repa}\) and denote
\(\boldsymbol{X}_j \equiv \begin{bmatrix} X_j^0 : \boldsymbol{X}_j^{*} \end{bmatrix}\)
as the basis function matrix for the \(j\)th variable after the
reparameterization. A spline function can be expressed in the matrix
form \[
B_j(x_j) = B_j^0(x_j) + B_j^*(x_j) = \beta_j X^0_j + \boldsymbol{\beta_j^*}^T \boldsymbol{X}_j^*,
\] and the generalized additive model in Equation (\ref{eq:gam}) now is
\begin{equation}\label{eq:gam-repa}
E(Y|\boldsymbol{x}) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} B_j(x_j)) = g^{-1}(\beta_0 + \sum\limits^p_{j=1} \boldsymbol{\beta}_j^T \boldsymbol{X}_j) = g^{-1}\left[\beta_0 + \sum\limits^p_{j=1} (\beta_j X^0_j + {\boldsymbol{\beta}_j^*}^T \boldsymbol{X}_j^*)\right],
\end{equation} where the coefficients
\(\boldsymbol{\beta}_j \equiv \begin{bmatrix} \beta_j : \boldsymbol{\beta}^*_j \end{bmatrix}\)
is an augmentation of the coefficient scalar \(\beta_j\) of linear space
and the coefficient vector \(\boldsymbol{\beta}^*_j\) of nonlinear
space.

@@

We introduce a novel SSL-based prior for smooth functions in GAMs. Given
the reparameterized design matrix
\(\boldsymbol{X}_j = \begin{bmatrix} X^0_j : \boldsymbol{X}_j^*\end{bmatrix}\)
for the \(j\)th variable, we impose a two-part SSL prior to the
coefficients
\(\boldsymbol{\beta}_j = \begin{bmatrix} \beta_j : \boldsymbol{\beta}_j^*\end{bmatrix}\).
Specifically, the linear space coefficient has an SSL prior and the
nonlinear space coefficients share a group SSL prior,
\begin{align}\label{eq:bham_ssl}
  \beta_{j} | \gamma_{j},s_0,s_1 &\sim DE(0,(1-\gamma_{j}) s_0 + \gamma_{j} s_1) \nonumber \\
  \beta^*_{jk} | \gamma^*_{j},s_0,s_1 &\overset{\text{iid}}{\sim}DE(0,(1-\gamma^*_{j}) s_0 + \gamma^*_{j} s_1), k=1,\dots, K_j
\end{align} where \(\gamma_{j}\in\{0,1\}\) and
\(\gamma^*_{j}\in \{0,1\}\) are two latent indicator variables,
indicating if the model includes the linear effect and the nonlinear
effect of the \(j\)th variable respectively. \(s_0\) and \(s_1\) are
scale parameters, assuming \(0 < s_0 < s_1\) and given. These scale
parameters \(s_0\) and \(s_1\) can be treated as tuning parameters and
optimized via cross-validation, discussed in Section \ref{sec:tune}.

@@

\textcolor{red}{To allow the shrinkage to self-adapt to the sparsity and smoothing pattern of the data, }we
further specify the parameter \(\theta_j\) follows a beta distribution
with given shape parameters \(a\) and \(b\), \[
\theta_j \sim Beta(a, b).
\] The beta distribution is a conjugate prior to the binomial
distribution and hence provides some computation convenience.
\textcolor{red}{Having a prior distribution of $\theta_j$ enables the proposed prior to inherit the selective shrinkage property and self-adaptivity \cite{Bai2021Review} from the classical SSL prior. In other words, when a smooth function is significant, the coefficients of the smooth function escape the overall shrinkage and produce a more accurate estimate, particularly in pair with the smoothing penalty implicitly addressed via the reparameterization. Meanwhile, the hyper prior encourages information borrowing across coordinates and hence automatic adjust for different levels of sparsity.}
Hereafter, we refer to the Bayesian hierarchical generalized additive
models with the two-part spike-and-slab LASSO prior as BHAM, and
visually presented in Figure \ref{fig:SSprior}.

@@

In this section, we compare the performance of the proposed model to
\textcolor{red}{six alternative models: linear LASSO models,} component
selection and smoothing operator (COSSO) \citep{Zhang2006GAM}, adaptive
COSSO \citep{Storlie2011}, generalized additive models with automatic
smoothing (referred to as \textit{mgcv} hereafter)\citep{Wood2011},
\textcolor{red}{spike-and-slab GAM \cite{Scheipl2012}}, and SB-GAM
\citep{Bai2021}.
\textcolor{red}{We use linear LASSO models as the benchmark, examining the performance when the linearity assumption doesn't hold}.
COSSO is one of the earliest smoothing spline models that consider
sparsity-smoothness penalty. Adaptive COSSO improved upon COSSO by using
adaptive weight for penalties such that the penalty of each functional
component is different for extra flexibility. \textit{mgcv} is one of
the most popular models for nonlinear effect interpolation and
prediction. Nevertheless,
\textcolor{red}{mgcv doesn't support analyses when the number of parameters exceeds the sample size.}
\textcolor{red}{Spike-and-slab GAM employs a spike-and-slab prior for GAM and uses an MCMC algorithm for model fitting.}
SB-GAM is the first spike-and-slab LASSO GAM. We implement linear LASSO
model with R package \texttt{glmnet 4.1-2}, COSSO and adaptive COSSO
with R package \texttt{cosso 2.1-1}, generalized additive models with
automatic smoothing with R package \texttt{mgcv 1.8-31}, spike-and-slab
GAM with R package \texttt{spikeSlabGAM 1.1-15}, and SB-GAM with R
package \texttt{sparseGAM 1.0}. COSSO models and SB-GAM do not provide
flexibility to define smooth functions, and hence use the default
choices; mgcv, spikeSlabGAM and the proposed model allow customized
smooth functions and we choose the cubic regression spline. We control
the dimensionality of each smooth function, 10 bases, for all different
choices of smooth functions. We use a 5-fold CV with the default
selection criteria to select the final model for linear LASSO model,
COSSO models, SB-GAM and the proposed model. Twenty default candidates
of tuning parameters (\(s_0\) in BHAM, \(\lambda_0\) in SB-GAM) are
examined for SB-GAM and the proposed model that allow user specification
of tuning candidates. All computation was conducted on a
high-performance 64-bit Linux platform with 48 cores of 2.70GHz
eight-core Intel Xeon E5-2680 processors and 24G of RAM per core and R
3.6.2 \citep{R}.

@@

\textcolor{red}{In the paper, we described a novel high-dimensional generalized additive model with Bayesian hierarchical prior for the purpose of predictive modeling. In particular, we introduce a two-part spike-and-slab LASSO prior for reparameterized smooth functions and derive a scalable EM-CD algorithm for model fitting. The proposed model provides a new angle to address the excess shrinkage of smooth functions that is commonly vulnerable to previous regularized high-dimensional GAMs, and hence improves the predictive performance. Th EM-CD algorithm, extended from previous spike-and-slab LASSO models, provides a computationally efficient alternative to the computational prohibitive MCMC algorithms, enhancing the scalability of spike-and-slab models. In addition, the two-part prior motivates the bi-level selection of predictors, selection of linear and nonlinear components. In the simulation study and real-data analyses, the proposed model demonstrates improvement in prediction and computational advantage when compared to the state-of-the-art models. When serving the purpose of variable selection, trade-offs exist among methods of comparison.}
We implement the proposed model in an open-source R package
\texttt{BHAM}, deposited at \url{https://github.com/boyiguo1/BHAM}. To
maximize the flexibility of smooth function specification, we deploy the
same programming grammar as in the state-of-the-art package
\texttt{mgcv}, in contrast to previous tools where smooth functions are
limited to the default ones. Ancillary functions are provided for model
specification in high-dimensional settings, curve plotting and
functional selection.

The proposed model shares many commonalities with the SB-GAM
\citep{Bai2021}, which is independently developed around the same time
as the proposed work. Both frameworks emphasize computational efficiency
by deploying group spike-and-slab LASSO type priors and
optimization-based scalable algorithms. Bai provides the theoretical
proof for the consistency of variable selection using group
spike-and-slab LASSO prior.
\textcolor{red}{The proposed model focuses on improving prediction performance for high-dimensional GAM, with the capability of bi-level selection. Moreover, the proposed model can easily generalize to other families of priors and smooth functions if desired. Not focused in this manuscript, the generalization is described in the Supporting Information.}
